A Standardized Effect Size for Evaluating the Comparing the Strength of Phylogenetic Signal

### 2020-07-16 Manuscript/OldVsns/Manuscript-3-EKB.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{setspace}\doublespacing
\usepackage{lineno}\linenumbers

@@

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with smaller values signifying `weak' phylogenetic signal,
while larger values are interpreted as `strong' phylogenetic signal
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Other approaches for interpreting \(\lambda\) are more statistical. For
instance, some have evaluated whether the observed \(\lambda\) differs
from some expected value through the use of confidence intervals
(Vandelook et al. 2019) or by performing likelihood ratio tests that
compare the observed model fit to that obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019). Additionally, qualitative comparisons of \(\lambda\) estimates
obtained from multiple phenotypic traits have been used to infer whether
the strength of phylogenetic signal is greater in one trait as compared
to another (e.g., Liu et al. 2019; Bai et al. 2019). Indeed, statements
regarding the strength of phylogenetic signal based on \(\lambda\) are
rather common in the evolutionary literature. For instance, of the 204
papers published in 2019 that estimated and reported Pagel's \(\lambda\)
(found in Google.scholar), 40\% interpreted the strength of phylogenetic
signal for at least one phenotypic trait. Additionally, because nearly
half of the 1,572 \(\lambda\) values reported were near the limits of
the parameter (Figure 1), this percentage is even higher, as biological
interpretation of phylogenetic signal at the limits of \(\lambda\) are
known. \hfill\break

@@

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. Specifically, \(\lambda=0\) represents no
phylogenetic signal, while \(\lambda=1\) is phylogenetic signal as
expected under Brownian motion. However, equating values of \(\lambda\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., the number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also assumes that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we still
lack a general understanding of the precision with which \(\lambda\) can
estimate levels of phylogenetic signal in phenotypic datasets.
\hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure the magnitude of a statistical effect in data, represented on a
common scale (Glass 1976; Cohen 1988). Effect sizes have widespread use
in many areas of the quantiative sciences, as they represent measures
that may be readily summarized across datasets as in meta-analysis
(Glass 1976; Hedges and Olkin 1985; Arnqvist and Wooster 1995), or
compared among datasets (e.g., Adams and Collyer 2016, 2019a).
Unfortunatley, not all model parameters and test statistics are effect
sizes, and thus many summary measures must first be converted to
standardized units (i.e., an effect size) for meaningful comparison (see
Rosenthal 1994). As a consequence, it follows that only if \(\lambda\)
is a statistical effect size can comparisons of estimates across
datasets be interpretable. For the case of \(\lambda\), this has not yet
been explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable estimate of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets, and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(Kappa\). Through simulations
across a wide range of conditions, we find that the precision of effect
sizes based on \(\lambda\) (\(Z_{\lambda}\)) are less reliable than that
those based on \(Kappa\) (\(Z_K\)), implying that \(Z_K\) is a more
robust effect size measure. We also propose a two-sample test statistic
that may be used to compare the strength of phylogenetic signal among
datasets, and provide an empirical example to demonstrate its use. We
conclude that estimates of phylogenetic signal using Pagel's \(\lambda\)
are often inaccurate, and thus interpreting strength of phylogenetic
signal in phenotypic datasets based on this measure is compromised. By
contrast, effect sizes obtained from \(Kappa\) hold promise for
characterizing phylogenetic signal, and for comparing the strength of
phylogenetic signal across datasets.

@@

We conducted a series of computer simulations to evaluate the precision
of Pagel's \(\lambda\). Our primary simulations were based on pure-birth
phylogenies; however, we also evaluated patterns on both balanced and
pectinate trees to determine whether tree shape affected our findings
(see Supporting Information). First we generated 50 pure-birth
phylogenies at each of six different tree sizes, ranging from 32 to 1024
taxa (\(n=2^5 - 2^{10}\)). Next, we rescaled the simulated phylogenies
by multiplying the internal branches by \(\lambda_{in}\), using 21
intervals of 0.05 units across its range
(\(\lambda_{in} = 0.0 \to 1.0\)), resulting in 1050 scaled phylogenies
at each level of species richness (\(n\)). Continuous traits were then
simulated on each phylogeny under a Brownian motion model of evolution
to obtain datasets with differing levels of phylogenetic signal, that
ranged from no phylogenetic signal (when \(\lambda_{in} =0\)), to
phylogenetic signal corresponding reflecting Brownian motion (when
\(\lambda_{in} =1\)). For each dataset we then estimated phylogenetic
signal (\(\lambda_{est}\)), and calculated the precision of \(\lambda\)
using the variance (\(\sigma^2_\lambda\)) across datasets at each input
level of phylogenetic signal and level of species richness. \hfill\break

We also evaluated the precision of \(\lambda\) when estimated in PGLS
regression and ANOVA (i.e., \(Y\sim{X}\)). Here, an independent variable
\(X\) was simulated on each phylogeny under a Brownian motion model of
evolution (for PGLS regression). For phylogenetic ANOVA, random groups
(\(X\)) were obtained by simulating a discrete (binary) character on
each phylogeny. Next, the dependent variable was simulated in such a
manner as to contain a known relationship with \(X\) plus random error
containing phylogenetic signal. This was accomplished as:
\(Y=\beta{X}+\epsilon\). Here, the association between \(Y\) and \(X\)
was modeled using a range of values:
\(\beta=(0.0,0.25, 0.5, 0.75,1.0)\), and the residual error was modeled
to contain phylogenetic signal simulated under a Brownian motion model
of evolution: \(\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})\): (see
Revell 2010 for a similar simulation design). The fit of the
phylogenetic regression was estimated using maximum likelihood, and
parameter estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were
obtained. Precision estimates (\(\sigma^2_\lambda\)) at each input level
of phylogenetic signal and level of species richness were then observed.
\hfill\break

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
obtained in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these empirical estimates were explicitly interpreted in terms
of the strength of phylogenetic signal that they represented (i.e.,
weak, intermediate, strong phylogenetic signal). \hfill\break

Finally, when \(\lambda\) was co-estimated with regression parameters in
PGLS regression, the results of our simulations were quite similar.
Here, regression parameters (\(\beta\)) were accurately estimated,
confirming earlier findings of Revell (2010) (see Supporting
Information). However, estimates of phylogenetic signal (\(\lambda\))
were less precise (Figure 3), and the spread of \(\lambda_{est}\) was
similar to that observed when \(\lambda\) was estimated for only the
dependent variable, as in Figure 2. Taken together, these findings
reveal that \(\lambda_{est}\) does not precisely characterize observed
levels of phylogenetic signal in phenotypic datasets, and that
biological interpretations of the strength of phylogenetic signal based
on \(\lambda\) may be highly inaccurate. \hfill\break

@@

To evaluate the utility of \(Z_K\) and \(Z_\lambda\) we calculated both
effect sizes for the simulated datasets generated above, and summarized
the precision of each using its variance (\(\sigma^2_{Z_K}\) and
\(\sigma^2_{Z_\lambda}\)). Results are found in Figure 4 (additional
results are found in the Supporting Information). Here two things are
evident. First, estimates of \(Z_K\) track the input phylogenetic signal
in a more linear fashion than do estimates of \(Z_\lambda\) (Figure
4A,B). Thus, actual changes in the strength of phylogenetic signal are
reflected more evenly in the corresponding values of the effect size
\(Z_K\). Second, the precision of \(Z_K\) is considerably more stable as
compared with \(Z_\lambda\). This may be seen by calculating the
coefficients of variation for the set of precision estimates (i.e.,
\(\sigma^2_{Z_K}\) and \(\sigma^2_{Z_\lambda}\)) across input levels of
phylogenetic signal. Here coefficients of variation in the precision of
\(Z_K\) were an order of magnitude smaller for than for \(Z_\lambda\)
(Figure 4C,D). This implied that estimates of the strength of
phylogenetic signal were more reliable and robust when using \(Z_K\) as
compared with \(Z_\lambda\). \hfill\break

@@

Once the magnitude of phylogenetic signal is characterized using
\(Z_K\), one may wish to compare such measures across datasets, to
determine whether the strength of phylogenetic signal is significantly
greater in one phenotypic trait as compared with another. As with other
effect sizes derived from permutation distributions (e.g., Adams and
Collyer 2016, 2019a), a two-sample test statistic may be found as:

@@

One important advantage of the approach advocated here is that the
resulting effect sizes (\(Z_K\)) are dimensionless, as the units of
measurement cancel out during the calculation of \(Z\) (Sokal and Rohlf
2012). As a consequence, the effect sizes represent the strength of
phylogenetic signal on a common and comparable scale, measured in
standard deviation units, regardless of the initial units and scale of
the original phenotypic variables under investigation. This means that
conceivably, the strength of phylogenetic signal may be compared across
datsets for phenotypic traits in different units and scale (e.g.,
behavioral and morphological traits), so long as they are continuous
variables. Note that this is directly related to the common practice of
converting data to their standard normal deviates prior to their use in
multivariate analysis, if the original variables are expressed in
different units or scales (see Everitt and Hothorn 2011; Legendre and
Legendre 2012; Adams and Collyer 2019b).

@@

Patterns worse with PGLS, though beta still estimated properly.
Conclusion, lambda not overly useful.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\usepackage[normalem]{ulem}
% Avoid problems with \sout in headers with hyperref
\pdfstringdefDisableCommands{\renewcommand{\sout}{}}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{setspace}\doublespacing
\usepackage{lineno}\linenumbers

@@

Evolutionary biologists commonly seek to describe the relative strength
of phylogenetic signal in phenotypic traits, to determine the extent to
which shared evolutionary history has influenced trait covariation among
taxa. This is often accomplished by interpreting empirical estimates of
\(\lambda\); with smaller values signifying `weak' phylogenetic signal,
while larger values are interpreted as `strong' phylogenetic signal
(e.g., De Meester et al. 2019; Pintanel et al. 2019; Su et al. 2019).
Other approaches for interpreting \(\lambda\) are more statistical. For
instance, some have evaluated whether the observed \(\lambda\) differs
from some expected value through the use of confidence intervals
(Vandelook et al. 2019) or by performing likelihood ratio tests that
compare the observed model fit to that obtained when \(\lambda=0\) or
\(\lambda=1\) (Freckleton et al. 2002; Cooper et al. 2010; Bose et al.
2019). Additionally, qualitative comparisons of \(\lambda\) estimates
obtained from multiple phenotypic traits have been used to infer whether
the strength of phylogenetic signal is greater in one trait as compared
to another (e.g., Liu et al. 2019; Bai et al. 2019). Indeed, statements
regarding the strength of phylogenetic signal based on \(\lambda\) are
rather common in the evolutionary literature; \sout{. For instance,} of
the 204 papers published in 2019 that estimated and reported Pagel's
\(\lambda\) (found in Google.scholar), 40\% EXPLICITLY interpreted the
strength of phylogenetic signal for at least one phenotypic trait.
\sout{Additionally, because nearly half of the 1,572 \(\lambda\) values
reported were near the limits of the parameter (Figure 1), this
percentage is even higher, as biological interpretation of phylogenetic
signal at the limits of \(\lambda\) are known.} FURTHERMORE, THIS
PERCENTAGE IS EVEN HIGHER WHEN CONSIDERING THE IMPLICIT BIOLOGICAL
INTERPRETATION OF PHYLOGENETIC SIGNAL FOR \(\lambda\)s NEAR 0 OR 1, AS
IS THE CASE FOR NEARLY HALF OF THE 1,572 REPORTED \(\lambda\) VALUES
(FIGURE 1). \hfill\break

@@

It seems intuitive to interpret the strength of phylogenetic signal
based on the value of \(\lambda\), as \(\lambda\) is a parameter on a
bounded scale (\(0\to1\)) for which interpretation of its extremal
points are understood. Specifically, \(\lambda=0\) represents no
phylogenetic signal, while \(\lambda=1\) is phylogenetic signal as
expected under Brownian motion. However, equating values of \(\lambda\)
directly to the strength of phylogenetic signal presumes two important
statistical properties that have not been fully explored. First, it
presumes that values of \(\lambda\) can be precisely estimated, as
biological inferences regarding the strength of phylogenetic signal
depend on high accuracy in its estimation. Therefore, understanding the
precision in estimating \(\lambda\) is paramount. One study (Boettiger
et al. 2012) found that estimates of Pagel's \(\lambda\) displayed less
variation (i.e., greater precision) when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)). From this
observation it was concluded that insufficient data (i.e., the number of
species) was the underlying cause of the increased variation across
parameter estimates (Boettiger et al. 2012). Indeed, such a pattern is
common with statistical estimators, as summary statistics and parameters
are often more precise at greater sample sizes (Cohen 1988). However,
this conclusion also assumes that the precision of \(\lambda\) remains
constant across its range (\(\lambda = 0 \to 1\)); an assumption that to
date, has not been verified. Thus, despite widespread use of Pagel's
(1999) \(\lambda\) in macroevolutionary studies, at present, we
\sout{still} lack a general understanding of the precision with which
\(\lambda\) can estimate levels of phylogenetic signal in phenotypic
datasets. \hfill\break

Second, while estimates of \(\lambda\) are within a bounded scale
(\(0\to1\)), this does not \emph{de-facto} imply that the estimated
values of this parameter correspond to the actual strength of the
underlying input signal in the data. For this to be the case,
\(\lambda\) must be a statistical effect size. Effect sizes are a
measure OF the magnitude of a statistical effect in data, represented on
a common scale (Glass 1976; Cohen 1988). Effect sizes have widespread
use in many areas of the quantiative sciences, as they represent
measures that may be readily summarized across datasets as in
\sout{meta-analysis} META-ANALYSES (Glass 1976; Hedges and Olkin 1985;
Arnqvist and Wooster 1995), or compared among datasets (e.g., Adams and
Collyer 2016, 2019a). \sout{Unfortunatley, not all model parameters and
test statistics are effect sizes, and thus many summary measures must
first be converted to standardized units (i.e., an effect size) for
meaningful comparison (see Rosenthal 1994). As a consequence,} THUS, it
follows that only \sout{if \(\lambda\) is a statistical effect size} BY
CONVERTING \(\lambda\) TO AN EFFECT SIZE can comparisons of estimates
across datasets be interpretable. For the case of \(\lambda\), this has
not yet been explored. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) for
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We find
that estimates of \(\lambda\) vary widely for a given input value of
phylogenetic signal, and that the precision in estimating \(\lambda\) is
not constant across its range. Rather, there is decreased precision when
input levels of phylogenetic signal are of intermediate strength.
Additionally, the same estimated values of \(\lambda\) may be obtained
from datasets containing vastly different input levels of phylogenetic
signal. Thus, \(\lambda\) is not a reliable estimate of the strength of
phylogenetic signal in phenotypic data. We then describe a standardized
effect size for measuring the strength of phylogenetic signal in
phenotypic datasets and apply the concept to two common measures of
phylogenetic signal: \(\lambda\) and \(Kappa\). Through simulations
across a wide range of conditions, we find that the precision of effect
sizes based on \(\lambda\) (\(Z_{\lambda}\)) are less reliable than
\sout{that} those based on \(Kappa\) (\(Z_K\)), implying that \(Z_K\) is
a more robust effect size measure. We also propose a two-sample test
statistic that may be used to compare the strength of phylogenetic
signal among datasets and provide an empirical example to demonstrate
its use. We conclude that estimates of phylogenetic signal using Pagel's
\(\lambda\) are often inaccurate, and thus interpreting strength of
phylogenetic signal in phenotypic datasets based on this measure is
compromised. By contrast, effect sizes obtained from \(Kappa\) hold
promise for characterizing phylogenetic signal, and for comparing the
strength of phylogenetic signal across datasets.

@@

We conducted a series of computer simulations to evaluate the precision
of Pagel's \(\lambda\). Our primary simulations were based on pure-birth
phylogenies; however, we also evaluated patterns on both balanced and
pectinate trees to determine whether tree shape affected our findings
(see Supporting Information). First we generated 50 pure-birth
phylogenies at each of six different tree sizes, ranging from 32 to 1024
taxa (\(n=2^5 - 2^{10}\)). Next, we rescaled the simulated phylogenies
by multiplying the internal branches by \(\lambda_{in}\), using 21
intervals of 0.05 units across its range
(\(\lambda_{in} = 0.0 \to 1.0\)), resulting in 1050 scaled phylogenies
at each level of species richness (\(n\)). Continuous traits were then
simulated on each phylogeny under a Brownian motion model of evolution
to obtain datasets with differing levels of phylogenetic signal, that
ranged from no phylogenetic signal (when \(\lambda_{in} =0\)), to
phylogenetic signal \sout{corresponding} reflecting Brownian motion
(when \(\lambda_{in} =1\)). For each dataset we then estimated
phylogenetic signal (\(\lambda_{est}\)), and calculated the
\sout{precision} VARIANCE of \(\lambda\) \sout{using the variance}
(\(\sigma^2_\lambda\)) across datasets at each input level of
phylogenetic signal and level of species richness AS A MEASURE OF
ESTIMATE PRECISION. \hfill\break

We also evaluated the precision of \(\lambda\) when estimated in PGLS
regression and ANOVA (i.e., \(Y\sim{X}\)). Here, an independent variable
\(X\) was simulated on each phylogeny under a Brownian motion model of
evolution (for PGLS regression). For phylogenetic ANOVA, random groups
(\(X\)) were obtained by simulating a discrete (binary) character on
each phylogeny. Next, the dependent variable was simulated in such a
manner as to contain a known relationship with \(X\) plus random error
containing phylogenetic signal. This was accomplished as:
\(Y=\beta{X}+\epsilon\). Here, the association between \(Y\) and \(X\)
was modeled using a range of values:
\(\beta=(0.0,0.25, 0.5, 0.75,1.0)\), and the residual error was modeled
to contain phylogenetic signal simulated under a Brownian motion model
of evolution: \(\epsilon=\mathcal{N}(\mu=0,\sigma=\mathbf{C})\): (see
Revell 2010 for a similar simulation design). The fit of the
phylogenetic regression was estimated using maximum likelihood, and
parameter estimates (\(\beta_{est}\) and \(\lambda_{est}\)) were
obtained. WE THEN CALCULATED Precision estimates (\(\sigma^2_\lambda\))
at each input level of phylogenetic signal and level of species
richness. \sout{were then observed.} \hfill\break

@@

\emph{Results.} We found that the precision of \(\lambda_{est}\) varied
widely across simulation conditions. Predictably, precision improved as
the number of species increased (Figure 2). This confirmed earlier
findings of Boettiger et al.~(2012), and adhered to parametric
statistical theory. However, in many cases the set of \(\lambda_{est}\)
spanned nearly the entire range of possible values (e.g., \(n=32\);
\(\lambda_{in}=0.5\): \(\lambda_{est}= 0.0\to 0.985\)), revealing that
estimates of \(\lambda\) were not a reliable indicator of input
phylogenetic signal. Importantly, the precision of \(\lambda_{est}\) was
not uniform across all levels of phylogenetic signal. The worst
precision was observed at intermediate levels of phylogenetic signal
(\(\lambda_{in}\approx0.5\)), while precision improved as input levels
approached the extremes of \(\lambda\)'s range (i.e.,
\(\lambda_{in}\rightarrow0\) \& \(\lambda_{in}\rightarrow1\)). Thus,
estimates of \(\lambda\) were least reflective of the true input signal
at intermediate values. Additionally, even at large levels of species
richness, we found that the range of \(\lambda_{est}\) still encompassed
a substantial portion of possible values (e.g., \(n=512\);
\(\lambda_{in}=0.5\): \(\lambda_{est} = 0.32\to 0.68\)). Likewise, the
same \(\lambda_{est}\) could be obtained from datasets containing vastly
different input levels of phylogenetic signal (e.g., \(n=512\);
\(\lambda_{est} = 0.5\); \(\lambda_{in} = 0.25 \to 0.65\)). These
findings were particularly unsettling when considered in light of our
literature survey. Over one quarter of the \(\lambda\) estimates
obtained in empirical studies (421 of 1,572) were between
\(\lambda=0.25\) and \(\lambda=0.75\) (Figure 1). This range reflected
the region that our simulations identified as being the least reliable
in terms of accurately characterizing levels of phylogenetic signal, yet
30\% of these MID-RANGE empirical estimates were explicitly interpreted
in terms of the strength of phylogenetic signal that they represented
(i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

Finally, when \(\lambda\) was co-estimated with regression parameters in
PGLS regression AND ANOVA, the results of our simulations were quite
similar. Here, regression parameters (\(\beta\)) were accurately
estimated, confirming earlier findings of Revell (2010) (see Supporting
Information). However, estimates of phylogenetic signal (\(\lambda\))
were less precise (Figure 3; FIGURE S1), and the spread of
\(\lambda_{est}\) was similar to that observed when \(\lambda\) was
estimated for only the dependent variable, as in Figure 2. Taken
together, these findings reveal that \(\lambda_{est}\) does not
precisely characterize observed levels of phylogenetic signal in
phenotypic datasets, and that biological interpretations of the strength
of phylogenetic signal based on \(\lambda\) may be highly inaccurate.
\hfill\break

@@

To evaluate the utility of \(Z_K\) and \(Z_\lambda\) we calculated both
effect sizes for the simulated datasets generated above, and summarized
the precision of each using its variance (\(\sigma^2_{Z_K}\) and
\(\sigma^2_{Z_\lambda}\)). Results are found in Figure 4 (additional
results are found in the Supporting Information). Here two things are
evident. First, estimates of \(Z_K\) track the input phylogenetic signal
in a more linear fashion than do estimates of \(Z_\lambda\) (Figure
4A,B). Thus, actual changes in the strength of phylogenetic signal are
reflected more evenly in the corresponding values of the effect size
\(Z_K\). Second, the precision of \(Z_K\) is considerably more stable as
compared with \(Z_\lambda\). This may be seen by calculating the
coefficients of variation for the set of precision estimates (i.e.,
\(\sigma^2_{Z_K}\) and \(\sigma^2_{Z_\lambda}\)) across input levels of
phylogenetic signal. Here coefficients of variation in the precision of
\(Z_K\) were an order of magnitude smaller for than for \(Z_\lambda\)
(Figure 4C\sout{,D}). This implied that estimates of the strength of
phylogenetic signal were more reliable and robust when using \(Z_K\) as
compared with \(Z_\lambda\). \hfill\break

@@

Once the magnitude of phylogenetic signal is characterized using
\(Z_K\), one may wish to compare such measures across datasets, to
determine whether the strength of phylogenetic signal is significantly
greater in one phenotypic trait as compared \sout{with} TO another. As
with other effect sizes derived from permutation distributions (e.g.,
Adams and Collyer 2016, 2019a), a two-sample test statistic may be
\sout{found} CALCULATED as:

@@

One important advantage of the approach advocated here is that the
resulting effect sizes (\(Z_K\)) are dimensionless, as the units of
measurement cancel out during the calculation of \(Z\) (Sokal and Rohlf
2012). As a consequence, the effect sizes represent the strength of
phylogenetic signal on a common and comparable scale, measured in
standard deviation units, regardless of the initial units and ORIGINAL
scale of the \sout{original} phenotypic variables under investigation.
This means that conceivably, the strength of phylogenetic signal may be
compared across datsets for phenotypic traits in different units and
scale (e.g., behavioral and morphological traits), so long as they are
continuous variables. Note that this is directly related to the common
practice of converting data to their standard normal deviates prior to
their use in multivariate analysis, if the original variables are
expressed in different units or scales (see Everitt and Hothorn 2011;
Legendre and Legendre 2012; Adams and Collyer 2019b).

@@

Patterns \sout{worse} STILL BAD with PGLS, though beta still estimated
properly. Conclusion, lambda not overly useful.
